### 2018-05-24 Datelife_paper1.Rmd

In the past decades, the possibility to obtain DNA sequences in large quantities from a wide variety of organisms became a reality, which, coupled to methodological development in phylogenetic and dating inference, allowed the application of molecular dating methods on a very large amount and diversity of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon ages across the tree of life.
To date, there is a large amount of both fossil and molecular-based data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE and Open Tree of Life (OToL).
OToL alone holds more than 200 chronograms. <!--cache updated jan 15 2018
devtools::load_all(pkg = "~/Desktop/datelife")
length(opentree_chronograms$trees)
names(opentree_chronograms)
-->
<!--How many trees? How many dated trees?-->
Methods to include living and fossil lineages are in continued development and increased usage by the community<!--total evidence and revbayes (fossilized birth-death)-->, which coupled to better sharing data practices, are greatly contributing to the accumulation in number and type of available data on taxon ages.

<!--What led to datelife development?
Describe public and research necessities covered by datelife
It is important to use available data on time frame of lineage origin:
To know the state of dating for a group of interest:
What range of estimated ages exist already? Are fossil and molecular
time frames coherent? (e.g., [@Magallon2015c]).
To construct a time tree of life.
For science communication, improve scientific discussions, time-framing
other events of importance in other research areas.
Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
-->
One inspiration for the DateLife project was the TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017]. This project has aggregated chronograms from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow. However, even in this gold standard resource, the included taxa only include between 0.097 and 3.236% of total species diversity (following taxonomic expert opinion on the global, extant species numbers, which ranges from 3 to 100 million species [Mayr2010; Moran2011]). One advantage of TimeTree is that it includes taxa from across the tree of life, versus more specialized chronograms focusing on plants [PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups. Users can use a web interface or a mobile app to receive information on divergence times for pairs of species, ______. It is of limited utility for scientific studies, however. The thousands of trees that have been entered are unavailable for examination or reuse; according to their FAQ, methods for allowing download of the data have been under discussion since 2015. There is no API allowing programmatic access to any data. The nearly hundred thousand taxon chronogram is not available with [@Kumar2017] or the TimeTree website, though the still substantial chronogram from @Hedges2015 is available. Large-scale, automated, data-mining is not allowed under the site's terms of use. As a science communication tool it is very powerful, since it has a friendly graphical interface, with informative while colorful outputs, that allows the general public to satisfy curiosity regarding their favorite organism or group of them.

>>>>

In the past decades, the possibility to obtain DNA sequences in large quantities from a wide variety of organisms became a reality, which, coupled to methodological development in phylogenetic and dating inference, allowed the application of molecular dating methods on a very large amount and diversity of organisms, greatly increasing the quantity of<!--organisms with--> data on taxon ages across the tree of life.
<!-- figure showing growth on studies with age data from references table from ttol-->
To date, there is a large amount of both fossil and molecular-based data on taxon ages and phylogenetic relationships in public repositories such as Dryad, TreeBASE and Open Tree of Life (OToL).
OToL alone holds more than 200 chronograms. <!--cache updated jan 15 2018
devtools::load_all(pkg = "~/Desktop/datelife")
length(opentree_chronograms$trees)
names(opentree_chronograms)
-->
<!--How many trees? How many dated trees?-->
Methods to include living and fossil lineages are in continued development and increased usage by the community<!--total evidence and revbayes (fossilized birth-death)-->, which coupled to better sharing data practices, are greatly contributing to the accumulation in number and type of available data on taxon ages.

<!--What led to datelife development?
Describe public and research necessities covered by datelife
It is important to use available data on time frame of lineage origin:
To know the state of dating for a group of interest:
What range of estimated ages exist already? Are fossil and molecular
time frames coherent? (e.g., [@Magallon2015c]).
To construct a time tree of life.
For science communication, improve scientific discussions, time-framing
other events of importance in other research areas.
Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
-->
One inspiration for the DateLife project was the TimeTree project [@Hedges2006; @Hedges2015; @Kumar2017]. This project has aggregated chronograms from 3,163 studies, encompassing 97,085 species [@Kumar2017], and continues to grow. However, even in this gold standard resource, the included taxa only include between 0.097 and 3.236% of total species diversity (following taxonomic expert opinion on the global, extant species numbers, which ranges from 3 to 100 million species [Mayr2010; Moran2011]). One advantage of TimeTree is that it includes taxa from across the tree of life, versus more specialized chronograms focusing on plants [PHYLOMATIC], birds [JETZ ET AL BIRDTREE.ORG], and other groups. Users can choose between a web interface or a mobile app to receive information on divergence times for the evolutionary history of a lineage, pairs of taxa, all lineages within a taxon, or a list of taxa. As a science communication tool, TimeTree project is a very powerful tool: it has a friendly graphical interface, with informative while colorful outputs, that allows the general public to satisfy curiosity regarding their favorite organism or group of them. It is of limited utility for scientific studies, however. The thousands of trees that have been entered are unavailable for examination or reuse; according to their FAQ, methods for allowing download of the data have been under discussion since 2015. Also, there is no Application Programming Interface (API) allowing programmatic access to any data. Moreover, large-scale, automated, data-mining is not allowed under the site's terms of use. The nearly hundred thousand taxon chronogram is not available with [@Kumar2017] or the TimeTree website, though the still substantial chronogram from [@Hedges2015] is available at OToL.
